Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2016-06-17 docs/appendixB.Rmd
adds new appendix; includes ms revisions; ready for coauthors

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

@@

Appendix B: Detailed methods for knot choice and placement
==========================================================
Our dimension reduction approach requires fitting random effects associated with knots distributed nonrandomly in space. By distributing *m* knots across space, we can fit random effects for each knot and then use process convolution to interpolate the random effect between the knots. Increasing the number of knots results in a more finely-resolved spatial effect, but it also increases processe time. Thus, there exists a tension between the resolution of the spatial effect and computational effort. Without regard to computation time, it is possible to use common model selection techniques to guide the number of knots one should use. With regard to computation time, it is sensible to choose knot placement and number based on the observed residual spatial variation.

To do so, we fit a generalized linear model of the form

\vspace{-4em}
\begin{align}
y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
\end{align}
\vspace{-4em}

where $\mu_{i,t}$ is the expected percent cover of pixel *i* in year *t*

\vspace{-4em}
\begin{align}
\text{log}(\mu_{i,t}) &= \beta_{0} + \beta_{1}y_{i,t-1} + \textbf{x}_{t}'\boldsymbol{\gamma}
\end{align}
\vspace{-3em}

that includes a density-dependence effect of log-transformed cover in the previous year ($y_{i,t-1}$) and climate effects ($\textbf{x}_t$). In other words, a non-hierarchial version of the model presented in our main text. We then examined a spatial variogram of the residuals from the above model to identify the the distance at which spatial dependence disappears.

Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig. B1). Thus, we set the kernal bandwidth ($\sigma$ in Eq. X of the main text) to 500/3 = `r round(500/3,2)`. We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters. Knots were placed evenly across space in the x and y directions. This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig. B2).

![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernel bandwidth in Eq. 6 of the main text (500/3; dashed black line).](../results/avgresidual_variogram_withrange_subset.png)

![Map of our study area showing percent cover in 1984 and the location of the 231 evenly-spaced knots.](../results/SAGE_Grid_wKnots_subset.png)

>>>>

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{graphicx}
bibliography: ~/Dropbox/Bibliography/SageAbundance.bib
csl: components/ecology.csl

@@

Appendix B: Annual maps of sagebrush cover
==========================================

We used remotely-sensed estimates of sagebrush cover [@Homer2012].
Here we show the annual maps of sagebrush cover as hindcasted with the regression tree algorithm described in @Homer2012.

\begin{figure}[!ht]
  \centering
      \includegraphics[width=7in]{../figures/all_years_percCover.png}
  \caption{Estimated percent cover of 30 x 30 meter pixels in our study area in southwestern Wyoming (see Fig. 1). Maps show percent cover for each year (1984-2011).}
\end{figure}

### 2016-06-17 docs/appendixC.Rmd
adds year effect to sims; updates docs with Mevin's comments
adds new appendix; includes ms revisions; ready for coauthors

Appendix C: Model convergence diagnostics
=========================================
Scale reduction factors [@Gelman1992; @Gelman2009] were less than 1.1 for all fitted parameters (Figure C1).
Visual inspection of traceplots indicates convergence (Figures C2-C10).

Table:  Definition of parameters and model notation for Figures C1-C10.

| Parameter | Definition | Notation in figures |
| --------- | ---------- | -------------------------------------- |
| $\alpha_{i}$ | spatial random effect for for knot *i* | ``alpha.i`` where ``i`` is the numeric knot id |
| $\gamma_{x}$ | climate effect for covariate *x* | ``beta.x`` where ``x`` is the numeric climate covariate id |
| $\beta_{0t}$ | random effect for year *t* on the intercept | ``int_yr.t`` where ``t`` is the numeric year id |
| $\bar{\beta_{0}}$ | mean intercept | ``int_mu`` |
| $\beta_{1}$ | density-dependent effect | ``beta_mu`` |
| $\sigma_{\beta_{0}}^2$ | variance of year random effect | ``sig_yr`` |
| $\sigma_{\eta}^2$ | variance of spatial random effect | ``sig_a`` |

![Scale reduction factors for each parameter ($\hat{R}$, points) and the unacceptable level of 1.1 shown in red.](../results/rhat_plot.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_1.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_2.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_3.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_4.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_5.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_6.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_7.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_8.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_9.png)

References
----------

>>>>

Appendix C: Detailed methods for knot choice and placement
==========================================================
Our dimension reduction approach requires fitting random effects associated with knots distributed nonrandomly in space. By distributing *m* knots across space, we can fit random effects for each knot and then use process convolution to interpolate the random effect between the knots. Increasing the number of knots results in a more finely-resolved spatial effect, but it also increases processe time. Thus, there exists a tension between the resolution of the spatial effect and computational effort. Without regard to computation time, it is possible to use common model selection techniques to guide the number of knots one should use. With regard to computation time, it is sensible to choose knot placement and number based on the observed residual spatial variation.

To do so, we fit a generalized linear model of the form

\vspace{-4em}
\begin{align}
y_{i,t} \sim \text{Poisson}(\mu_{i,t}),
\end{align}
\vspace{-4em}

where $\mu_{i,t}$ is the expected percent cover of pixel *i* in year *t*

\vspace{-4em}
\begin{align}
\text{log}(\mu_{i,t}) &= \beta_{0} + \beta_{1}y_{i,t-1} + \textbf{x}_{t}'\boldsymbol{\gamma}
\end{align}
\vspace{-3em}

that includes a density-dependence effect of log-transformed cover in the previous year ($y_{i,t-1}$) and climate effects ($\textbf{x}_t$). In other words, a non-hierarchial version of the model presented in our main text. We then examined a spatial variogram of the residuals from the above model to identify the the distance at which spatial dependence disappears.

Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig. C1). Thus, we set the kernel bandwidth ($\sigma$ in Eq. 6 of the main text) to 500/3 = `r round(500/3,2)` because the effective spatial range under an exponential covariance structure is equal to $\sigma \times 3$. We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters. Knots were placed evenly across space in the x and y directions. This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig. C2).

![Spatial variogram of model residuals showing where spatial dependence disappears (~500 meters; asymptote of the variogram), and the value chosen for the kernel bandwidth in Eq. 6 of the main text (500/3; dashed black line).](../figures/avgresidual_variogram_withrange_subset.png)

![Map of our study area showing percent cover in 1984 and the location of the 231 evenly-spaced knots.](../figures/SAGE_Grid_wKnots_subset.png)

### 2016-06-17 docs/appendixD.Rmd
adds new appendix; includes ms revisions; ready for coauthors

\renewcommand\thefigure{D\arabic{figure}}

Appendix D: Posterior estimate of spatial random effects ($\boldsymbol{\eta}$)
==============================================================================

```{r figure_D1,dependson="plot-options", fig.cap="Standardized percent cover showing the general spatial structure of above and below average pixels in the data (top panel) compared to the mean posterior estimate of the spatial field, $\\eta$ (bottom panel).", eval=TRUE, echo=FALSE}
library(png)
library(grid)
img <- readPNG("../results/spatialfield.png")
grid.raster(img)
```

>>>>

\renewcommand\thefigure{D\arabic{figure}}  
\renewcommand\thetable{D\arabic{table}}

Appendix D: Model convergence diagnostics
=========================================
Scale reduction factors [@Gelman1992; @Gelman2009] were less than 1.1 for all fitted parameters (Figure D1).
Visual inspection of traceplots indicated convergence.

\footnotesize

Table:  Definition of parameters and model notation for Figure D1.

| Parameter | Definition | Notation in figures |
| --------- | ---------- | -------------------------------------- |
| $\alpha_{i}$ | spatial random effect for for knot *i* | ``alpha.i`` where ``i`` is the numeric knot id |
| $\gamma_{x}$ | climate effect for covariate *x* | ``beta.x`` where ``x`` is the numeric climate covariate id |
| $\beta_{0t}$ | random effect for year *t* on the intercept | ``int_yr.t`` where ``t`` is the numeric year id |
| $\bar{\beta_{0}}$ | mean intercept | ``int_mu`` |
| $\beta_{1}$ | density-dependent effect | ``beta_mu`` |
| $\sigma_{\beta_{0}}^2$ | variance of year random effect | ``sig_yr`` |
| $\sigma_{\eta}^2$ | variance of spatial random effect | ``sig_a`` |

\normalsize

![Scale reduction factors for each parameter ($\hat{R}$, points) and the unacceptable level of 1.1 shown in red.](../figures/rhat_plot.png)

<!--

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_01.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_02.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_03.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_04.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_05.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_06.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_07.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_08.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_09.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_10.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_11.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_12.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_13.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_14.png)

![Traceplots of three MCMC chains for each model parameter.](../results/poissonSage_randYear_traceplots_Page_15.png)

-->

References
----------

### 2016-06-17 docs/ms_sageAbundance.Rmd
adds new appendix; includes ms revisions; ready for coauthors

Materials and Methods
---------------------
### Data
#### Remotely-sensed time series
To demonstrate our modeling approach, we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
As part of a separate study, \hl{Homer et al. 2012 estimated} sagebrush percent cover using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
The regression tree model was further validated using another 297 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images \hl{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}. 
This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Fig. A1).
In this \hl{remote sensing product}, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
Thus, the full \hl{remote sensing product} contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).

@@

and $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
It is possible, through exhaustive model selection and fitting, to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig. B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
An infinite number of knots would result in an exact representation of the spatial process and covariance model.
Computationally, using an infinite number of knots is not possible, thus the use of dimension reduction techniques serves as an approximation, where the accuracy increases with the number of knots.
Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.

@@

### Fitting the model
We fit the spatiotemporal model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Short chains of samples are a hallmark of the Stan algorithm, which is extremely efficient.
Compared to other samplers, fewer iterations are required to achieve convergence.
Each chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days on a four node Central Processing Unit with 2 $\times$ AMD Opteron(tm) Processor 4386 @ 3.10 Ghz, 64GB of RAM per node, 16 cores per node, and each chain launched in parallel on separate cores.
We assessed convergence visually and calculated scale-reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].

@@

All parameters in our model converged on stable posterior distributions (Appendix C).
Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change (Fig. 3).
In other words, if the year 2000 water year was wetter than average, sagebrush cover would increase from the 2001 to the 2002 growing season.
Other climate effects strongly overlapped zero but their posterior means were positive, except for fall-through-spring precipitation the first year of a cover transition (*t*-1), whose posterior mean was negative (Fig. 3).
The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. D1).
This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.

@@

When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5 and Fig. E1).
For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig. 6).
Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change estimated for our study area (Fig. 3).

Discussion
-----------
Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
We \hl{introduced} a new approach \hl{that combines a} remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling \hl{literature} [e.g., @Conn2015].
\hl{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming.}
We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.

@@

Under projected climate, we forecast modest increases in sagebrush cover for all RCP scenarios (Figs. 5,6).
Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
Cumulative precipitation the year before a given year-to-year transition was the strongest standardized effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2) and the negative effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*) had an offsetting effect.
In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).

An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecasted increase in spatial heterogeity (Fig. E1).
Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). 
Had we not explicitly accounted for spatial-dependence in our model, we would have missed this result.
We were unable to attribute the spatial structure apparent in the data (Fig. 4A) and approximated by our model ($\boldsymbol{\eta}$, Fig. D1) to slope, aspect, elevation, or coarse soil type (results not shown).
The lack of correlation between $\boldsymbol{\eta}$ and landscape factors leads us to conclude that the spatial structure in our data set emerges from some combination of fine-scale microhabitat associations and legacy effects related to fire and species invasions.

While we forecast an increase in sagebrush cover at our study area, SDM studies typically project dramatic declines in climate suitability for sagebrush with warming [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
There are many potential explanations for this apparent contrast, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in a cold portion of sagebrush's geographic distribution may be the best.
The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt because their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
The average annual temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 was 4.6\textdegree C (calculated from Daymet estimates).
Our study area lies at the cold extreme of the sagebrush range, thus the weak positive response to temperature that we estimated (Fig. 3) and carried through to our forecasts (Figs. 5,6) likely represents the positive effect of earlier snowmelt, and thus higher moisture availability early in the growing season.

@@

We projected sagebrush cover to the end of this century, but an important feature of our approach is that it could also produce short-term forecasts.
For example, we could forecast the effects of a multi-year regional drought on sagebrush cover [@Debinski2010]. Validating spatial population models against short-term predictions would give ecological forecasters a way to assess and improve the performance of their models, which would greatly increase our confidence in long-term forecasts.
This cycle of prediction, validation, and refinement is missing from most currently available population-level forecasts of the effects of climate change.

>>>>

Materials and Methods
---------------------
### Data
#### Remotely-sensed time series
To demonstrate our modeling approach, we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
As part of a separate study, \hl{Homer et al. 2012 estimated} sagebrush percent cover using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
The regression tree model was further validated using another 297 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46% [@Homer2012].
To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images \hl{to generate yearly predictions of sagebrush cover for all of Wyoming for the years 1984-2011}. 
This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011 (Fig. B1).
In this \hl{remote sensing product}, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
Thus, the full \hl{remote sensing product} contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).

@@

and $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
It is possible, through exhaustive model selection and fitting, to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig. C2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix C).
An infinite number of knots would result in an exact representation of the spatial process and covariance model.
Computationally, using an infinite number of knots is not possible, thus the use of dimension reduction techniques serves as an approximation, where the accuracy increases with the number of knots.
Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.

@@

### Fitting the model
We fit the spatiotemporal model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Short chains of samples are a hallmark of the Stan algorithm, which is extremely efficient.
Compared to other samplers, fewer iterations are required to achieve convergence.
Each chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days on a four node Central Processing Unit with 2 $\times$ AMD Opteron(tm) Processor 4386 @ 3.10 Ghz, 64GB of RAM per node, 16 cores per node, and each chain launched in parallel on separate cores.
We assessed convergence visually and calculated scale-reduction factors (Appendix D, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].

@@

All parameters in our model converged on stable posterior distributions (Appendix D).
Only the *lagPpt* climate covariate can be considered important based on a 90% credible interval, and it had a positive effect on sagebrush percent cover change (Fig. 3).
In other words, if the year 2000 water year was wetter than average, sagebrush cover would increase from the 2001 to the 2002 growing season.
Other climate effects strongly overlapped zero but their posterior means were positive, except for fall-through-spring precipitation the first year of a cover transition (*t*-1), whose posterior mean was negative (Fig. 3).
The posterior mean for the spatial random effect, $\boldsymbol{\eta}$, captured the overall spatial structure of the observed data (Fig. E1).
This indicates our choice of knot placement and dimension reduction strategy was adequate for describing permanent spatial variation in the data.

@@

When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6A).
The spatial effect remains strong enough in low cover regions to counteract the positive effect of projected precipitation increases (Fig. 5).
Thus, our model predicts an increase in the heterogeneity of sagebrush cover because projected cover increases are smaller in low cover pixels than in high cover pixels (Fig. 5 and Fig. F1).
For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig. 6A).
Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6A).
The generally increasing trend reflects the positive effect of precipitation on sagebrush cover change estimated for our study area (Fig. 3).
We also show how our model is capable of near-term forecasts in Fig 6B.

Discussion
-----------
Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
We \hl{introduced} a new approach \hl{that uses methods from the dynamic spatio-temporal modeling literature} [e.g., @Conn2015] \hl{to fit a population model to remotely-sensed estimated of plant percent cover}.
\hl{As a proof-of-concept, we applied our approach to a remotely-sensed data product of sagebrush percent cover from 1984 to 2011 in Wyoming} [@Homer2012].
We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.

@@

Under projected climate, we forecast modest increases in sagebrush cover for all RCP scenarios in the long-term (Figs. 5,6A).
Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
Cumulative precipitation the year before a given year-to-year transition was the strongest standardized effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2) and the negative effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*) had an offsetting effect.
In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).

An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecasted increase in spatial heterogeity (Fig. F1).
Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). 
Had we not explicitly accounted for spatial-dependence in our model, we would have missed this result.
We were unable to attribute the spatial structure apparent in the data (Fig. 4A) and approximated by our model ($\boldsymbol{\eta}$, Fig. E1) to slope, aspect, elevation, or coarse soil type (results not shown).
The lack of correlation between $\boldsymbol{\eta}$ and landscape factors leads us to conclude that the spatial structure in our data set emerges from some combination of fine-scale microhabitat associations and legacy effects related to fire and species invasions.

While we forecast an increase in sagebrush cover at our study area, SDM studies typically project dramatic declines in climate suitability for sagebrush with warming [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
There are many potential explanations for this apparent contrast, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in a cold portion of sagebrush's geographic distribution may be the best.
The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt because their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
The average annual temperature across the sagebrush steppe biome is 6.9\textdegree C [sd = 1.6; @Schlaepfer2011], whereas average temperature at our study area from 1980 to 2013 was 4.6\textdegree C (calculated from Daymet estimates).
Our study area lies at the cold extreme of the sagebrush range, thus the weak positive response to temperature that we estimated (Fig. 3) and carried through to our forecasts (Figs. 5,6A) likely represents the positive effect of earlier snowmelt, and thus higher moisture availability early in the growing season.

@@

We projected sagebrush cover to the end of this century, but an important feature of our approach is that it can also produce short-term forecasts (Fig. 6B).
For example, we could forecast the effects of a multi-year regional drought on sagebrush cover [@Debinski2010]. Validating spatial population models against short-term predictions would give ecological forecasters a way to assess and improve the performance of their models, which would greatly increase our confidence in long-term forecasts.
This cycle of prediction, validation, and refinement is missing from most currently available population-level forecasts of the effects of climate change.
